---
title: "NEWS-MBSStools"
output: github_document
  #md_document:
   # variant: markdown_github
---

<!-- NEWS.md is generated from NEWS.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "NEWS-"
)
```

```{r, echo = FALSE}
cat(paste0("Last Update: ",Sys.time()))
```

Version history.

# Planned Updates

* Each tool to be a separate version.

* PHI; need to update included data.

# v0.0.0.9002
2017-04-20

* Flow sum calculation completed.  Included Channel Number as grouping variable but sum across channels to get the total site flow.

* PHI, edit script for outlining in RStudio.

* PHI.calc() to PHIcalc().

* Data.R and folder "data-raw".

* PHI, tweak for created data.  Using Excel to build rda rather than txt file from the Excel.

# v0.0.0.9001
2017-03-29

* Completed PHI calculation.


# v0.0.0.9000
2017-03-27

* Created GitHub repository.

* Added ReadME.RMD and NEWS.RMD

* PHI calculation started.
