---
title: "NEWS-MBSStools"
output: github_document
  #md_document:
   # variant: markdown_github
---

<!-- NEWS.md is generated from NEWS.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "NEWS-"
)
```

```{r, echo = FALSE}
cat(paste0("Last Update: ",Sys.time()))
```

Version history.

# Planned Updates

* None at this time.

# v1.0.2.9003
2017-07-30

* Add continuous integration (TravisCI) badge to ReadMe.

# v1.0.2.9002
2017-07-25

* DESCRIPTION
* Move dplyr and rgdal from Suggests to Imports
    + Packages not always loading.  Issue #15 and Issue #17
* Modify minimum version of R from 3.3.2 to 3.4.0.  
    + MapTaxaObs did not work in earlier version of R.  
    Updated to v3.4.0 and no issue.  Issue #17.

# v1.0.2.9001
2017-05-11

* Issue #17 with MapTaxaObs
* Removed extra file in vignette folder.
* Update Vignette as doesn't build in check; commented out MapTaxaObs.

# v1.0.2.0000
2017-10-26

* Corrected metric.score() [line 181] for fish for "HIGHLANDS" was incorrectly designated.  These sites were not getting a final FIBI score.

# v1.0.1.0000
2017-09-26

* Update each function's description and vignette with the packages needed for that function to work properly.

* Update Vignette for a few typos and so shows in final package.  Also messages=FALSE for XLConnect example.

* DESCRIPTION.  Move rgdal from Depends to Suggests.

* Fish metric example not working in metric.values().  Issue #14.  Extra parameter.  Removed and works.

# v1.0.0.0000
2017-09-24

* Release version 1.

# v0.0.0.9018
2017-09-24

* Add install guide notebook.

* Tweak vignette.

* Update Readme.

* Final version before going to v1.0.0.

* IonContrib, ensure data input is a data frame.

* PHI, ensure data input is a data frame.

# v0.0.0.9017
2017-09-20

* Update vignette.  

* Fixed example for fish metrics in metric.values().  20170920.

# v0.0.0.9016
2017-09-19

* IonContrib() added barplot example.  20170919.

# v0.0.0.9015
2017-09-13

* Fix typo in example for metric.scores(); "Metrics.Bugs.Scores" to "Metrics.Bugs.Scores.MBSS". 20170913.

# v0.0.0.9014
2017-09-09

* Added MSW (family) data, values, and scores.  20170909.

# v0.0.0.9013
2017-09-09

* Added BIBI.
+ Updated metric names to match those used by MBSS.
+ Updated metric.values() and metric.scores().
+ Updated genus level bug data file.

# v0.0.0.9012
2017-09-08

* Add FIBI.  2017-09-05.
+ metric.values.R
+ metric.score.R
+ example data files
+ \data-raw\metric.scoring.tab and metric adjustment
* Added dplyr to DESCRIPTION.  Needed for metric values and scores.  20170908
* Modified \data-raw\metrics.scoring.tab to resolve QC issues with 2016 data.  20170908.

# v0.0.0.9011
2017-08-22

* Updated taxa map data for typo in "no fish observed".  20170822.

# v0.0.0.9010
2017-07-31

* Updated fish data and crosswalk table examples so have more matched data.  Also update help to remind that data has to be on the first worksheet for the Excel file.  20170731.

# v0.0.0.9009
2017-07-28

* Updated DESCRIPTION.  "Erik" to "Erik W." and added Matt Ashton as data contributor.  20170728.

# v0.0.0.9008
2017-07-21

* Added TaxaMapsObs().  Issue #2. 20170720.

* Libraries readxl (suggested) and rgdal (required) added to DESCRIPTION.  Maps won't generate without rgdal loaded. 20170720.

* Added example data and GIS files (zipped). 20170721.

# v0.0.0.9007
2017-06-23

* Added "LazyData: true" to DESCRIPTION.  Data was not being loaded properly.  Issue #9.  20170623.

# v0.0.0.9006
2017-06-23

* Modified vignette.

* Modified data-raw Process data R files (wd).

* Added IonContrib.R

# v0.0.0.9005
2017-06-14

* Added vignette for documentation of each function and the library as a whole.

# v0.0.0.9004
2017-04-28 (not complete)

* Address issues with PHIcalc from testing.  Removed extra rows from Excel import for data MBSS.PHI.  Added line to convert values to numeric inside MBSS.PHI.

# v0.0.0.9003
2017-04-21

* FlowSum().  Added Error Handling for misidentified Channels.  Updated data MBSS.flow.

# v0.0.0.9002
2017-04-20

* Flow sum calculation completed.  Included Channel Number as grouping variable but sum across channels to get the total site flow.

* PHI, edit script for outlining in RStudio.

* PHI.calc() to PHIcalc().

* Data.R and folder "data-raw".

* PHI, tweak for created data.  Using Excel to build rda rather than txt file from the Excel.

# v0.0.0.9001
2017-03-29

* Completed PHI calculation.


# v0.0.0.9000
2017-03-27

* Created GitHub repository.

* Added ReadME.RMD and NEWS.RMD

* PHI calculation started.
