---
title: "MBSStools Vignette"
author: "Erik Leppo"
date: "2017-06-14"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MBSStools Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction
MBSStools was created by Tetra Tech in 2017 to meet the needs of Maryland Department of Natural Resources (DNR), Maryland Biological Stream Survey (MBSS) staff to complete data manipulations in an efficient and reproducible manner.

# Install

The R library is saved on GitHub for ease of updating and distribution.  Issues can be tracked and then fixed.  Current users can update the library using the same code used to install the library (see below).

```{r, eval=FALSE}
devtools::install_github("leppott/MBSStools")
```

To contact the author directly email Erik.Leppo@tetratech.com.

# Contents
Main functions:

1. Index of Biological Integrity (IBI) calculations for fish and benthic macroinvertebrates.

2. Taxonomic distribution maps for fish.

3. Stream discharge calculation.

4. Ion contribution to conductivity.

5. Physical Habitat Index (PHI) calculation.

Example data included with each function.

## IBI

MBSStools calculates both the benthic macroinvertebrate IBI (BIBI) and fish IBI (FIBI).

### IBI, Benthic Macroinvertebrates
```{r, eval=FALSE}
#example code
```
### IBI, Fish
```{r, eval=FALSE}
#example code
```
## Fish Distribution Maps

## Stream Discharge
```{r, eval=FALSE}
# data
data("MBSS.flow")
# calculate flow
flow.cell <- FlowSum(MBSS.flow,returnType="cell")
flow.sample <- FlowSum(MBSS.flow)
# examine data
View(flow.cell)
View(flow.sample)
# Example Save file (tab delimited text file)
datetime <- format(Sys.time(),"%Y%m%d_%H%M%S")
myYear <- "15"
write.table(flow.sample,paste0("SumFlow",myYear,"_",datetime,".tab"),row.names=FALSE,sep="\t")
```
## Ion Matrix
Data will need to be in "wide" format.  That is, a single record for each sample and the ions to be analyzed in columns.
```{r, eval=FALSE} 
#echo=FALSE, results='asis'
#knitr::kable(MBSS.Ion.Ref)
```

The columns for the ions need to match the reference table names.  The function IonContrib() includes a reference table that is used by default but the user can supply their own reference table.

```{r, eval=FALSE} 
#echo=FALSE, results='asis'
#knitr::kable(head(MBSS.Ion.Data))
```

Example
```{r, eval=FALSE}
# Load Data
data.ion <- MBSS.Ion.Data
# Load Reference Table
ref.ion <- MBSS.Ion.Ref
# Calculate Percent Contribution by Ion
contrib.ion <- IonContrib(data.calc, ref.ion)
View(contrib.ion)
# Save Results
#write.table
# Plot Results
#dotchart()
```

Dot charts are a good way to display the data.  * user has to decide needs for plotting *

* show example plots


## PHI
Based on document.
Provide link.

Calculates the proper pre-/post-2000 index.

Requirements

1. StationID

1. BioRegion

1. All other data fields.


data
```{r, echo=FALSE, results='asis'}
x <- data("MBSS.PHI")
knitr::kable(head(x))
```

Loading example data and performing calculation.  The results will have to be saved.
```{r, eval=FALSE}
# data
data("MBSS.PHI")
# calculate PHI
PHI <- PHIcalc(MBSS.PHI)
# examine data
View(PHI)
# Example Save file
datetime <- format(Sys.time(),"%Y%m%d_%H%M%S")
write.table(PHI,paste0("PHI_",datetime,".tab"),row.names=FALSE,sep="\t")
```

Result
```{r, echo=FALSE, results='asis'}
# data
#data(MBSS.PHI)
# calculate PHI
#PHI <- PHIcalc(MBSS.PHI)
#knitr::kable(head(PHI))
```
# Vignette Text

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))



